# Exploring the phylogenetic composition of microbiomes

Some of this code as been adapted from the [ECOSCOPE GitHub account](https://github.com/EDUCE-UBC) 

## Table of contents

#### 1. [Prior to the workshop](#Prior)

#### 2. [Introduction](#Introduction)

#### 3. [Building a microbial phylogenetic tree](#Building-a-microbial-phylogenetic-tree)

#### 4. [Diversity analysis in R ](#Diversity-analysis-in-R)

## 1. Prior to the workshop <a name="Prior"></a>

### Setup 

Please come to the workshop with your laptop set up with the required
software and data files as described in our [setup instructions](https://github.com/FloMazel/Microbiome_Phylo_Diversity_Workshop/blob/master/SetUp.md).

### Background 

Please read [Hallam SJ *et al*. 2017. Sci Data 4: 170158](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5663219/) "Monitoring microbial responses to ocean deoxygenation in a model oxygen minimum zone" to learn more about the data used in this workshop. You can also check out this [short video](https://drive.google.com/file/d/1s6V263-Vj2KQ6rExOBaYn-UDOERyQVGZ/view?usp=sharing) showing how the sampling was done!


### Data description
The data in this workshop were collected as part of an on-going oceanographic time series program in Saanich Inlet, a seasonally anoxic fjord on the East coast of Vancouver Island, British Columbia 


## 2. Introduction <a name="Introduction"></a>

### 2.1. Theory
General concepts: Alpha, Beta and Gamma diversity
Microbial units and the need of a phylogenetic framework

### 2.2. Getting started with R 

#### Installing and loading packages

At the beginning of every R script, you should have a dedicated space for loading R packages. R packages allow any R user to code reproducible functions and share them with the R community. Packages exist for anything ranging from microbial ecology to complex graphics to multivariate modeling and beyond. 

In this workshop, we will use several packages 
XXX

```{r, message=FALSE}
if (!require("seqinr")) install.packages("seqinr")
if (!require("ape")) install.packages("ape")
if (!require("vegan")) install.packages("vegan")
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("phyloseq")) install.packages("phyloseq")
if (!require("betapart")) install.packages("betapart")
if (!require("abind")) install.packages("abind")
if (!require("Matrix")) install.packages("Matrix")
if (!require("phytools")) install.packages("phytools")
if (!require("ggplot2")) install.packages("ggplot2")
```

#### Setting your working directory 

Copy the workshop folder "Working directory" from GitHub on your computer

Then tell R that this is going to be the folder where we are going to work 

In my case: 
```{r, message=FALSE}
setwd("/Users/jonsanders/Dropbox/Documents/Research/Education/Workshops/Microbiome_Phylo_Diversity_Workshop")
```

#### Importing custom R functions

We have written some custom R functions for use in this workshop. It is good
practice to define these functions in a separate file and then to import them
into your scripts, so that you don't duplicate the code. (Duplication, you 
may recall, can end up in mutation! ;) )

```{r, message=FALSE}
source("./R functions/BDTT_functions.R")
```

## 3. Building a microbial phylogenetic tree <a name="Building-a-microbial-phylogenetic-tree"></a>

### 3.1. Alignment <a name="Alignment"></a>

We are not going to cover this stage here, as it will be carried out in other workshops. 
Many software offer to align sequences, some of them are wrapped within [Mothur](https://www.mothur.org) or [Qiime2](https://qiime2.org)
For example you can look at this [tutorial](https://github.com/FloMazel/Microbiome_Phylo_Diversity_Workshop/blob/master/data/mothur_pipeline.html) producing the data we are going to use here. 

### 3.2. Tree Building <a name="Tree-Building"></a>

Typical read length in microbiome studies are relatively short and have thus contains limited information to reconstruct phylogeentic trees, especially to reconstruct deep branches. To avoid biased phylogenies, we thus constrains deep branches to follow taxonomic classification as it is admitted that large taxonomic clades are monophyletic (informations based on longer sequences). The choice of the constrains is not easy. Here we will constain tree reconstruction by Domain (Bacteria/Archea) and phylums.

We will use FastTree to reconsttuct the phylogeneic hypotheses. While fastTree is not the best software to reconstruct phylogenies (because it makes a lot of approximations), it has the huge avantage to be fast, which is often critical in microbial species, as there are a very high number of species. 

#### Topological constrains

Load the taxonomic file

```{r, message=FALSE}
taxonomy.raw = read.table("data/Saanich_cruise72_mothur_OTU_taxonomy.taxonomy", sep="\t", header=TRUE, row.names=1)
taxonomy= taxonomy.raw %>% 
  select(-Size) %>% 
  separate(Taxonomy, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep=";")
```

Change the name of the fasta alignment file 

```{r, message=FALSE}
alignment=read.fasta("data/mothur_intermediate_files/Saanich.final.opti_mcc.unique_list.0.03.rep.fasta")
names(alignment)=rownames(taxonomy)
```

Remove OTUS not assigned to a domain

```{r, message=FALSE}
taxonomy=subset(taxonomy,!Domain=="unknown")
```

Domain constrains

```{r, message=FALSE}
taxonomy[["Bacteria"]][taxonomy$Domain=="Bacteria"]=1
taxonomy[["Bacteria"]][taxonomy$Domain=="Archaea"]=0
```

Phylum constrains

```{r, message=FALSE}
Phylum=unique(taxonomy$Phylum)
Phylum=subset(Phylum,!Phylum=="unknown_unclassified") #remove this factor

for (i in Phylum)
  {
  taxonomy[[as.character(i)]][taxonomy$Phylum==i]=1
  taxonomy[[as.character(i)]][!taxonomy$Phylum==i]=0
  }

Constrains=taxonomy[,c("Bacteria",as.character(Phylum))] #keep only the constrains 
```

Convert to fasta file

```{r, message=FALSE}
sequences=list()
for (i in 1:dim(Constrains)[1]){sequences[[i]]=Constrains[i,]}
write.fasta(sequences, names=rownames(Constrains), file.out="My_outputs/Phylogenetic_Constrains.fasta", open = "w", nbchar = 60, as.string = FALSE)
```

Prune the alignment to sequences with assigned domains

```{r, message=FALSE}
alignment=alignment[rownames(Constrains)]
write.fasta(alignment, names=names(alignment), file.out="My_outputs/Saanich.final.opti_mcc.unique_list.0.03.rep_Names_Modified.fasta", open = "w", nbchar = 60, as.string = FALSE)
```


#### FastTree

Now we are going to use the alignment of the reads to construct a phylogenetic tree of the reads.

To do so we are going to use the FastTree software, that you should already have installed on your computer. 

On mac, you need to open a terminal and acces the software commands where it is located (in my computer here: "Desktop/Programmes_Unix/FastTree")

The FastTree function requires input on (1) the model used (here gtr + cat 20), (2) the evenutal topological constrains (hre the file "Phylogenetic_Constrains.fasta" we produces before in R), (3) the alignment (here "Saanich.final.opti_mcc.unique_list.0.03.rep_Names_Modified.fasta") and a name for the output tree (here we used "Saanish_FastTree"). You need to give the exact adress of the different files in the command line  (here my files are loacted in "Desktop/Recherche/En_cours/workshopMicrobiome/Contenu/MyFiles/"). For further help on the the funciton, you can type: 

```{r, message=FALSE}
Desktop/Programmes_Unix/FastTree -h
```

Now we run fastTree on our data:

```{r, message=FALSE}
Desktop/Programmes_Unix/FastTree -gtr -cat 20 -constraints Desktop/Recherche/En_cours/workshopMicrobiome/Contenu/MyFiles/Phylogenetic_Constrains.fasta -nt Desktop/Recherche/En_cours/workshopMicrobiome/Contenu/MyFiles/Saanich.final.opti_mcc.unique_list.0.03.rep_Names_Modified.fasta > Desktop/Recherche/En_cours/workshopMicrobiome/Contenu/MyFiles/Saanish_FastTree
```

For the sake of comparison, we will laso construct the tree WITHOUT topological constrains. In this case just remove the parameters "constrains" 

```{r, message=FALSE}
#Desktop/Programmes_Unix/FastTree -gtr -cat 20 -nt Desktop/Recherche/En_cours/workshopMicrobiome/Contenu/MyFiles/Saanich.final.opti_mcc.unique_list.0.03.rep_Names_Modified.fasta > Desktop/Recherche/En_cours/workshopMicrobiome/Contenu/MyFiles/Saanish_FastTree_withoutConstrains
```

### 3.3. Tree Visualisation <a name="Tree-Visualisation"></a>

We now are coming back to R. We first need to load the trees we just constructed 

```{r, message=FALSE}
Tree=read.tree('My_outputs/Saanish_FastTree')
TreeNoC=read.tree('My_outputs/Saanish_FastTree_withoutConstrains')
```

We can now plot the tree, with colours for Domains. First creating a palette of colours for the Domains and assign each OTU a color based on that. 

```{r, message=FALSE}
Domains=unique(taxonomy$Domain)
paletteDomains=rainbow(length(Domains));names(paletteDomains)=Domains 
coloursDomains=paletteDomains[as.character(taxonomy$Domain)];names(coloursDomains)=rownames(taxonomy)
```

Then plot and write the tree in pdf format. 

```{r, message=FALSE}
pdf("My_outputs/Phylogenetic_tree_colouredby_Domains.pdf",width=15,height=15)
plot(Tree,type="fan",cex=.3,tip.color=coloursDomains[Tree$tip.label])
legend(1, 0, legend=names(paletteDomains),fill=paletteDomains, cex=2)
dev.off()
```

We can do the same for phylum instead of domain. 

```{r, message=FALSE}
Phylums=unique(taxonomy$Phylum)
palettePhylums=rainbow(length(Phylums));names(palettePhylums)=Phylums #define colours for taxonomic groups (Phylums here)
coloursPhylums=palettePhylums[as.character(taxonomy$Phylum)];names(coloursPhylums)=rownames(taxonomy) #assign colors to each OTU depending on its taxonomic group 
pdf("My_outputs/Phylogenetic_tree_colouredby_Phylum.pdf",width=15,height=15)
plot(Tree,type="fan",cex=.3,tip.color=coloursPhylums[Tree$tip.label])
legend(1, 1, legend=names(palettePhylums),fill=palettePhylums, cex=1)
dev.off()
```

It would be now interesting to check the impact of the topological contrains on the trees. To do so, we simply plot the tree obtained without constrains and observe the distribution of phylum in this tree. 

```{r, message=FALSE}
pdf("My_outputs/Phylogenetic_tree_colouredby_Phylum_NoConstrains.pdf",width=15,height=15)
plot(TreeNoC,type="fan",cex=.3,tip.color=coloursPhylums[TreeNoC$tip.label])
legend(1, 1, legend=names(palettePhylums),fill=palettePhylums, cex=1)
dev.off()
```

Rapidly compare the two trees, what do you observe? 

### 3.3. Re-root the tree (Archaea Vs Bacteria) <a name="Tree-Rerooting"></a>

A phylogenetic tree is characterized by a root, which indicate the direction of evolution (from the root to the tip). For "prokariotes", we will assume to be on the branch separating bacteria from archaea. 

To do so, we first need to find this branch

```{r, message=FALSE}
Archaea=row.names(taxonomy)[taxonomy$Domain=="Archaea"]
MRCAnode=getMRCA(phy = Tree,tip = Archaea)
```

Then, we re-root the tree on this branch

```{r, message=FALSE}
TreeNewRoot=root(phy=Tree, node=MRCAnode,resolve.root = T)
```

and write the corresponding tree

```{r, message=FALSE}
write.tree(TreeNewRoot,'My_outputs/Saanish_FastTreeRooted')
```

We finally plot this tree and save the image. 

```{r, message=FALSE}
pdf("My_outputs/Phylogenetic_tree_colouredby_Domains_NewRoot.pdf",width=15,height=15)
plot(TreeNewRoot,type="fan",cex=.3,tip.color=coloursDomains[Tree$tip.label])
legend(1, 0, legend=names(paletteDomains),fill=paletteDomains, cex=2)
dev.off()
```

We now have a better tree. And we are going to use it to analyse microbiome composition patterns!

## 4. Diversity analysis in R <a name="Diversity-analysis-in-R"></a>

### 4.1. Basics <a name="Basics"></a>

Importing the data: OTU table, metadata and phylogenetic tree

The tree: 
```{r, message=FALSE}
Tree=read.tree('My_outputs/Saanish_FastTreeRooted')
```

The OTU table:
```{r, message=FALSE}
OTU = read.table("data/Saanich_cruise72_mothur_OTU_table.shared", sep="\t", header=TRUE, row.names=2)
OTU.clean = OTU %>%
  select(-label, -numOtus)
```

The taxonomic file:
```{r, message=FALSE}
taxonomy = read.table("data/Saanich_cruise72_mothur_OTU_taxonomy.taxonomy", sep="\t", header=TRUE, row.names=1)
taxonomy.clean = taxonomy %>% 
  select(-Size) %>% 
  separate(Taxonomy, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep=";")
```

The metadata:
```{r, message=FALSE}
metadata = read.table("data/Saanich_cruise72_metadata.txt", sep="\t", header=TRUE, row.names=1)
metadata=metadata[,c("Depth_m","PO4_uM","SiO2_uM","NO3_uM","NH4_uM","CH4_nM" ,"Salinity_PSU")]
```

We then Construct the phyloseq object:

First, we define parts of the phyloseq object.

```{r, message=FALSE}
OTU.clean.physeq = otu_table(as.matrix(OTU.clean), taxa_are_rows=FALSE)
tax.clean.physeq = tax_table(as.matrix(taxonomy.clean))
metadata.physeq = sample_data(metadata)
phylogeny.physeq=phy_tree(Tree)
```

and then assemble them

```{r, message=FALSE}
mothur = phyloseq(OTU.clean.physeq, tax.clean.physeq, metadata.physeq,phylogeny.physeq) 
mothur
```


### 4.2. Classic beta-diversity analysis <a name="Classics"></a>

We first compute classic metrics of beta-diversity

```{r, message=FALSE}
BC=vegdist(otu_table(mothur),method = "bray")
Jaccard=vegdist(otu_table(mothur),method = "jac")
UniFracBeta=UniFrac(mothur)
UniFracWBeta=UniFrac(mothur,weighted = T)
```

We can then plot them to see how much correlated they are
```{r, message=FALSE}
plot(Jaccard,UniFracBeta)
```

And then visualize the pattern of microbiome composition in relation to the metadata, for Unifrac:

```{r, message=FALSE}
ordi = ordinate(mothur, "PCoA", "unifrac", weighted=F)
plot_ordination(mothur, ordi, color="Depth_m")
```

or Bray Curtis: 

```{r, message=FALSE}
ordi = ordinate(mothur, "PCoA", "bray", weighted=F)
plot_ordination(mothur, ordi, color="Depth_m")
```

Importantly, we should also test the statistical relationship between microbiome beta-diversity and the metadata, using for example a PERMANOVA test: 

```{r, message=FALSE}
adonis(BC~Depth_m,data=data.frame(sample_data(mothur)))
adonis(Jaccard~Depth_m,data=data.frame(sample_data(mothur)))
adonis(UniFracBeta~Depth_m,data=data.frame(sample_data(mothur)))
```

### 4.3. Screening the phylogenetic scale <a name="BDTT"></a>

#### 4.3.1 Computing Beta-diversity


We now go a step ahead and screen the phylogentic scale to study the compositipon of the microbiomes. First, we check what are the values of this scale, i.e. the scale of divergence times, in substitution/site. 

```{r, message=FALSE}
Hnodes=getHnodes(Tree) 
hist(Hnodes,n=150,xlim=c(0,.5))
```

We can then define the set of discrete threshold used to build different set of microbial OTUs

```{r, message=FALSE}
slices=c(seq(from=0,to=0.3,by=0.025)) 
```

The idea the is to compute beta-doversity matrices for all these slices: 

We first extract a OTU table matrix from the phyloseq object
```{r, message=FALSE}
mat=t(as(otu_table(mothur), "matrix"))
```

and then run the analysis (for Jaccard and Bray Curtis)

```{r, message=FALSE}
MultipleBetaJac=BDTT(similarity_slices=slices,tree=Tree,sampleOTUs=mat,onlyBeta=T,metric="jac")
saveRDS(MultipleBetaJac,"My_outputs/Multiple_Resolution_Beta_Jaccard.RDS")  

MultipleBetaBC=BDTT(similarity_slices=slices,tree=Tree,sampleOTUs=mat,onlyBeta=T,metric="bc")
saveRDS(MultipleBetaJac,"My_outputs/Multiple_Resolution_Beta_BrayCurtis.RDS")  
```

#### 4.3.2. Statistical link to metadata

As for the simple beta-diversity tests, we then link the microbiome composition to our metadata. In our case, we need to construct models for each threshold independently:

We first prepare the result table: 

```{r, message=FALSE}
predictors=names(sample_data(mothur))
StatsRes=expand.grid(similarity_slices=as.character(slices),predictors=predictors,metric=c("Jac","BC"))
StatsRes[["F.Model"]]=StatsRes[["R2"]]=StatsRes[["Pr(>F)"]]=NA
```
and then "fill" it without the models that we construct in loop: 

```{r, message=FALSE}
for (i in as.character(slices))
{
  for (j in predictors) 
  {
   res=unlist(adonis(formula = MultipleBetaJac[i,,]~data.frame(sample_data(mothur))[,j])$aov.tab[1,c(4,5,6)])
   StatsRes[(StatsRes$metric=="Jac")&(StatsRes$predictors==j)&(StatsRes$similarity_slices==i),4:6]=res
   res=unlist(adonis(formula = MultipleBetaBC[i,,]~data.frame(sample_data(mothur))[,j])$aov.tab[1,c(4,5,6)])
   StatsRes[(StatsRes$metric=="BC")&(StatsRes$predictors==j)&(StatsRes$similarity_slices==i),4:6]=res
   }
}
```

We can then plot the profiles of R2 along the phylogenetic time scale:

```{r, message=FALSE}
ggplot(aes(y=R2,x=similarity_slices,colour=predictors,group=factor(predictors)),data=StatsRes)+geom_point()+geom_line()+facet_wrap(~metric)
ggsave("My_outputs/BDTT_Jaccard_BC.pdf",height = 7,width = 10)
```

### 4.5. Screening all the nodes of the phylogeny: PhyloFactor and Balance Trees <a name="screening-all-the-nodes"></a>

#### 4.5.1. PhyloFactor

```{r, message=FALSE}
require(phylofactor)
```

#### Loading data

We will load data the same data as above. PhyloFactor has some specific requirements for data format that will require further modifications, which are explained below.

The tree: 
```{r, message=FALSE}
Tree=read.tree('My_outputs/Saanish_FastTreeRooted')
```

The OTU table:
```{r, message=FALSE}
OTU = read.table("data/Saanich_cruise72_mothur_OTU_table.shared", sep="\t", header=TRUE, row.names=2)
OTU.clean = OTU %>%
  select(-label, -numOtus)

# phylofactor uses OTU tables in matrix format, with features as rows and samples as columns

OTU.matrix <- as.matrix(t(OTU.clean))

# we will need to subset the OTU table to only those tips found in the tree
OTU.matrix <- OTU.matrix[row.names(OTU.matrix) %in% Tree$tip.label,]
```

The taxonomic file:
```{r, message=FALSE}
taxonomy = read.table("data/Saanich_cruise72_mothur_OTU_taxonomy.taxonomy", sep="\t", header=TRUE, row.names=1)
taxonomy.clean = taxonomy %>% 
  select(-Size) %>% 
  separate(Taxonomy, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep=";")

taxonomy <- taxonomy %>% select(-Size)
taxonomy$OTU_ID = row.names(taxonomy)
taxonomy <- taxonomy[rownames(OTU.matrix),]
```

The metadata:
```{r, message=FALSE}
metadata = read.table("data/Saanich_cruise72_metadata.txt", sep="\t", header=TRUE, row.names=1)
metadata=metadata[,c("Depth_m","PO4_uM","SiO2_uM","NO3_uM","NH4_uM","CH4_nM" ,"Salinity_PSU")]

# similarly, we need to subset metadata to just those samples in our data

metadata <- metadata[colnames(OTU.matrix),]
```


#### Simulate factor on bacterial phylogeny

Fundamentally, PhyloFactorization is a method for seeing how characteristics change across a phylogeny, and identifying the edges of the phylogeny where they change most significantly. 

We can demonstrate this by simulating some random data on top of our previously calculated microbial phylogeny, and adjusting the values for a few of the groups. You can imagine this simulated data as a characteristic of the microbes -- say, genome size, or cell diameter. 

Here, we're creating a new variable called `CellDiam`, and for each tip on our tree `Tree`, drawing a value from the log normal distribution using the function `rlnorm`. Then, we'll re-draw values for different subsets of the tree using values that are increased or decreased by 4-fold. 

```{r}

set.seed(1)
CellDiam <- rlnorm(length(Tree$tip.label))

# let's make Arcobacter and Actinobacteria different from the rest

arcobacter <- Tree$tip.label %in% row.names(taxonomy.clean[taxonomy.clean$Genus == 'Arcobacter',])
actinobacteria <- (Tree$tip.label %in% row.names(taxonomy.clean[taxonomy.clean$Phylum == 'Actinobacteria',]))

CellDiam[arcobacter] <- rlnorm(sum(arcobacter))*4
CellDiam[actinobacteria] <- rlnorm(sum(actinobacteria))/4

logCellDiam <- log(CellDiam)
```

Now we can PhyloFactor the tree to discover the branches that lead to the tips with the changed values. The function `twoSampleFactor` uses two-sample tests to identify edges separating two parts of the tree that differ with respect to a data vector -- in this case, the simulated `CellDiam` values.

```{r}
pf_twoSample <- twoSampleFactor(logCellDiam, Tree, nfactors=5)
```

We can look at a summary of the results by calling the object:

```{r}
pf_twoSample
```

Note that it returned 5 factors, but we only changed two clades on the tree! What happened?

PhyloFactor works by sequentially partitioning the tree from the most to the least important groups with respect to your given criteria. Note that the first two factors do in fact correspond to large groups, roughly matching the clades we adjusted! The remaining top five factors are individual tips that randomly obtained extreme draws from the simulated distribution.

```{r}
print(sum(actinobacteria))
print(sum(arcobacter))
```

And although statistical significance is still not well-understood for PhyloFactorization, you can see from the vector of p-values associated with these factors that there is a steep drop-off after the first two factors:

```{r}
pf_twoSample$pvals
```

We can also visualize these phylogenetic factors on the phylogeny itself using the `pf.tree` function in PhyloFactor, which creates a `ggtree` plot highlighting each returned factor:

```{r}
cellDiam.tree <- pf.tree(pf_twoSample)
cellDiam.tree$ggplot
```

Factor colors are written out as a separate legend element in the returned object:

```{r}
cellDiam.tree$legend
```

#### Testing for phylogenetic groups correlated with metadata: linear regression

This will find the phylogenetic factors that vary most predictably with depth. By using the built-in choice function 'F', we are asserting that we want only the factors with the largest F score to be retained -- up to 9 factors. The `stop.early` parameter tells PhyloFactor to stop searching for new factors once it drops below a certain significance threshold (see the documentation for more on the heuristic used and its justification).


```{r}
pf_depth_F <- PhyloFactor(OTU.matrix, Tree, X = metadata, frmla = Data~Depth_m, nfactors = 9, ncores=2, stop.early = T, choice='F')
```


We can then do the same for the next factor.

```{r}
s_pf_depth.summary.2 <- summary(pf_depth_F, factor=2)

s_pf_depth.summary.2
```


In this case, the factor actually comprises 19 tips in a single monophyletic clade.

```{r}
pf.taxa(pf_depth_F, taxonomy, factor=2)$group1
```

Which, it turns out, encompasses bacteria in the Rhodobacteraceae group of the Alphaproteobacteria. This factor also appears to be tightly correlated to depth:

```{r}
ggplot(s_pf_depth.summary.2$data, aes(x = Depth_m, y = Data)) + 
geom_point() + 
geom_smooth(method='lm')
```


The previous results were ranked according to the factors that had the strongest correlation, as measured by effect size (F). We can also rank factors according to the total proportion of variance explained, by setting `choice` equal to `'var'`:

```{r}
pf_depth_var <- PhyloFactor(OTU.matrix, Tree, X = metadata, frmla = Data~Depth_m, nfactors = 9, ncores=2, stop.early = T, choice='var')
```

```{r}
pf_depth_var
```

Note that when optimizing for total variance explained, PhyloFactor will tend to find larger clades, as they represent a larger overall proportion of the data. 

```{r}
s_pf_depth.var.summary.1 <- summary(pf_depth_var, factor=1)
```

```{r}
pf.taxa(pf_depth_var, taxonomy, factor=1)$group1
```

Which, it turns out, encompasses bacteria in the Rhodobacteraceae group of the Alphaproteobacteria. This factor also appears to be tightly correlated to depth:

```{r}
ggplot(s_pf_depth.var.summary.1$data, aes(x = Depth_m, y = Data)) + 
geom_point() + 
geom_smooth(method='lm')
```

```{r}
s_pf_depth.var.summary.2 <- summary(pf_depth_var, factor=2)

pf.taxa(pf_depth_var, taxonomy, factor=2)$group1
```

```{r}
ggplot(s_pf_depth.var.summary.2$data, aes(x = Depth_m, y = Data)) + 
geom_point() + 
geom_smooth(method='lm')
```


#### Comparing phylofactorization with two groups

Let's combine all we've practiced so far to visualize using PhyloFactorization how different clades of the bacterial phylogeny respond to changes in different variables in our dataset. 

We can use the `cowplot` package to combine different ggplot elements to simultaneously illustrate the ILR-values for individual PhyloFactors and their location on the phylogeny. For this, we'll just take the top three factors 

First, we'll write a short helper function to format the individual inset XY plots.

```{r}

inset_plot <- function(pf, factor, variable, color) {
  pf_s <- summary(pf, factor=factor)
  
  p <- ggplot(pf_s$data, aes_string(x = variable, y = 'Data')) + 
        geom_point(color = color) + 
        geom_abline(slope = coef(pf$models[[factor]])[2],
                    intercept = coef(pf$models[[factor]])[1])
  
  return(p)
}
```

Now we can perform a phylofactorization of the bacterial phylogeny with respect to different variables of interest. We'll compare the bacterial groups responding to nitrate (NO3) and phosphate (PO4), two important and frequently limiting nutrients in oligotrophic environments. Let's grab the top three factors for each, according to total variance explained.

```{r}
pf_NO3_var <- PhyloFactor(OTU.matrix, Tree, X = metadata, frmla = Data~NO3_uM, nfactors = 3, ncores=2, choice='var')

pf_PO4_var <- PhyloFactor(OTU.matrix, Tree, X = metadata, frmla = Data~PO4_uM, nfactors = 3, ncores=2, choice='var')
```

For each PhyloFactor object, we can use the `cowplot` library to combine separate `ggplot2` panels into a publication-ready image, showing both individual factors locations on the tree, and their response (in ILR-transformed space) to each predictor variable.

```{r}
# Draw the PhyloFactors on the phylogeny
NO3_tree = pf.tree(pf_NO3_var)

# Make inset plots for each factor
g1_NO3 = inset_plot(pf_NO3_var,                  # The PhyloFactor object we just calculated
                    1,                           # The first factor
                    'NO3_uM',                    # Plotting against the NO3_uM variable
                    NO3_tree$legend$colors[1])   # Use the color for this factor from the tree
g2_NO3 = inset_plot(pf_NO3_var, 2, 'NO3_uM', NO3_tree$legend$colors[2])
g3_NO3 = inset_plot(pf_NO3_var, 3, 'NO3_uM', NO3_tree$legend$colors[3])

# Draw the inset plots composed with the PhyloFactor phylogeny
ggdraw() +
  draw_plot(NO3_tree$ggplot, 
            x = 0, y = 0.05, width = .7, height = .7,scale = 1.8) +
  draw_plot(g1_NO3, x = .7, y = .66, width = .3, height = .33) +
  draw_plot(g2_NO3, x = .7, y = .33, width = .3, height = .33) +
  draw_plot(g3_NO3, x = .7, y = 0, width = .3, height = .33) +
  draw_plot_label(label = c("Phylogeny", "A", "B","C"), size = 15,
                  x = c(0, 0.67, .67,.67), y = c(1, 1, 0.66,0.33))

```
```{r}
# Repeat for the phosphate PhyloFactors
PO4_tree = pf.tree(pf_PO4_var)

g1_PO4 = inset_plot(pf_PO4_var, 1, 'PO4_uM', PO4_tree$legend$colors[1])
g2_PO4 = inset_plot(pf_PO4_var, 2, 'PO4_uM', PO4_tree$legend$colors[2])
g3_PO4 = inset_plot(pf_PO4_var, 3, 'PO4_uM', PO4_tree$legend$colors[3])

ggdraw() +
  draw_plot(PO4_tree$ggplot, 
            x = 0, y = 0.05, width = .7, height = .7,scale = 1.8) +
  draw_plot(g1_PO4, x = .7, y = .66, width = .3, height = .33) +
  draw_plot(g2_PO4, x = .7, y = .33, width = .3, height = .33) +
  draw_plot(g3_PO4, x = .7, y = 0, width = .3, height = .33) +
  draw_plot_label(label = c("Phylogeny", "A", "B","C"), size = 15,
                  x = c(0, 0.67, .67,.67), y = c(1, 1, 0.66,0.33))

```
